Federated Learning's Blessing:\\ FedAvg has Linear Speedup

### 2021-07-05 source/sec_exp.tex
Update exp

@@

In this section, we empirically examine the linear speedup convergence of FedAvg and Nesterov accelerated FedAvg on various scenarios, including strongly convex function, convex smooth function, and overparameterized setting, as proved in previous theorems.

\textbf{Setup.} Follow the experimental setting in~\cite{stich2018local}, we
conduct experiments on both synthetic datasets and real-world dataset
w8a~\cite{platt1998fast} $(d=300, n=49749)$. We consider the distributed
objectives $F(\vw) = \sum_{k=1}^N p_kF_k(\vw)$, and the objective function on
$k$-th local device includes three cases: 1) \textbf{Strongly convex
objective}: the regularized binary logistic regression problem, $F_k(\vw) =
\frac{1}{N_k} \sum_{i=1}^{N_k} \log( 1+ \exp(-y_i^k \vw^T\vx_i^k) + \frac{\lambda}{2}
\|\vw\|^2$. The regularization parameter is set to $\lambda = 1/n \approx
2e-5$. 2) \textbf{Convex smooth objective}: the binary logistic regression
problem without regularization. 3) \textbf{Overparameterized setting}:
the linear regression problem without adding noise to the label, $F_k(\vw) =
\frac{1}{N_k} \sum_{i=1}^{N_k} (\vw^T\vx_i^k + b  - y_i^k)^2$.  
% In the first
% two cases, we use the w8a dataset to conduct empirical evaluation. 
% In the overparameterized setting, we use features from w8a dataset and 
% generate a ground truth $w$ to label each instance, which satisfies the
% overparameterized setting.

% ~\ref{th:strongcvxsmth}~\ref{th:cvxsmoth}~\ref{th:fedmasslr}
\textbf{Linear speedup of FedAvg and Nesterov accelerated FedAvg.} To verify the linear speedup convergence as shown in Theorem~\ref{thm:SGD_scvx}~\ref{thm:SGD_cvx}~\ref{thm:nesterov_scvx}~\ref{thm:Nesterov_cvx}, we evaluate the number of iterations needed to reach
$\epsilon$-accuracy in three objectives. We initialize all runs with $\vw_0 = \textbf{0}_d$ and measure the number of iterations to reach the target accuracy $\epsilon$. For each configuration $(E, K)$, we extensively search the learning rate from $\min(\eta_0, \frac{nc}{1 + t})$, where
$\eta_0 \in \{0.1, 0.12, 1, 32 \}$ according to different problems and $c$ can
take the values $c = 2^i \ \forall i \in \ZZ$. As the results shown in Figure~\ref{fig:speedup},
the number of iterations decreases as the number of (active) workers increasing, which is consistent for FedAvg and Nesterov accelerated FedAvg across all scenarios.
For the additional experiments on the impact of $E$, detailed experimental setup, and hyperparameter setting, please refer to the Appendix, section~\ref{sec:expsupp}.

@@

\begin{figure}[t!]
\centering
{\small
\begin{tabular}{ccc}
\hspace{-2em}\includegraphics[width=0.33\textwidth]{fig/paper-stronglycvxsmthspeedupNodesT-min-w8a-epsilon0131-reg1e-05.pdf} &
\includegraphics[width=0.33\textwidth]{fig/paper-cvxsmoothspeedupNodesT-min-w8a-epsilon0134-reg0.pdf} &
\includegraphics[width=0.33\textwidth]{fig/paper-linregressionspeedupNodesT-min-linearregressionw8a-epsilon002-reg0.pdf}\\
\hspace{-2em}\includegraphics[width=0.33\textwidth]{fig/paper-partialstronglycvxsmthspeedupNodesT-min-w8a-epsilon0131-reg1e-05.pdf} &
\includegraphics[width=0.33\textwidth]{fig/paper-partialcvxsmoothspeedupNodesT-min-w8a-epsilon0134-reg0.pdf} &
\includegraphics[width=0.33\textwidth]{fig/paper-partiallinregressionspeedupNodesT-min-linearregressionw8a-epsilon002-reg0.pdf}\\
\hspace{-2em}\includegraphics[width=0.33\textwidth]{fig/paper-nesterovspeedupNodesT-min-w8a-epsilon0131-reg1e-05.pdf} & 
\includegraphics[width=0.33\textwidth]{fig/paper-nesterovspeedupNodesT-min-w8a-epsilon0134-reg0.pdf}
& 
\includegraphics[width=0.33\textwidth]{fig/paper-lrnesterovspeedupNodesT-min-linearregressionw8a-epsilon002-reg0.pdf}\\
(a) Strongly convex objective & (b) Convex smooth objective & (c) Linear regression
	\end{tabular}}
	\vspace{-1em}
\caption{The linear speedup of FedAvg in full participation, partial participation, and the linear speedup of Nesterov accelerated FedAvg, respectively.
% \textbf{Linear speedup } The first row reports the linear speedup convergence of FedAvg w.r.t the number of workers. The second row reports the linear speedup convergence of FedAvg w.r.t the number of active workers in the partial parcipation setting. The third row reports the linear speedup convergence of Nesterov accelerated FedAvg in full participation setting. 
}
\vspace{-1em}
\label{fig:speedup}
\end{figure}

>>>>

\begin{figure*}[ht!]
\centering
{\small
\begin{tabular}{ccc}
\hspace{-2em}\includegraphics[width=0.33\textwidth]{fig/paper-stronglycvxsmthspeedupNodesT-min-w8a-epsilon0131-reg1e-05.pdf} &
\includegraphics[width=0.33\textwidth]{fig/paper-cvxsmoothspeedupNodesT-min-w8a-epsilon0134-reg0.pdf} &
\includegraphics[width=0.33\textwidth]{fig/paper-linregressionspeedupNodesT-min-linearregressionw8a-epsilon002-reg0.pdf}\\
\hspace{-2em}\includegraphics[width=0.33\textwidth]{fig/paper-partialstronglycvxsmthspeedupNodesT-min-w8a-epsilon0131-reg1e-05.pdf} &
\includegraphics[width=0.33\textwidth]{fig/paper-partialcvxsmoothspeedupNodesT-min-w8a-epsilon0134-reg0.pdf} &
\includegraphics[width=0.33\textwidth]{fig/paper-partiallinregressionspeedupNodesT-min-linearregressionw8a-epsilon002-reg0.pdf}\\
\hspace{-2em}\includegraphics[width=0.33\textwidth]{fig/paper-nesterovspeedupNodesT-min-w8a-epsilon0131-reg1e-05.pdf} & 
\includegraphics[width=0.33\textwidth]{fig/paper-nesterovspeedupNodesT-min-w8a-epsilon0134-reg0.pdf}
& 
\includegraphics[width=0.33\textwidth]{fig/paper-lrnesterovspeedupNodesT-min-linearregressionw8a-epsilon002-reg0.pdf}\\
(a) Strongly convex objective & (b) Convex smooth objective & (c) Linear regression
	\end{tabular}}
	\vspace{-1em}
\caption{The linear speedup of FedAvg in full participation, partial participation, and the linear speedup of Nesterov accelerated FedAvg, respectively.
% \textbf{Linear speedup } The first row reports the linear speedup convergence of FedAvg w.r.t the number of workers. The second row reports the linear speedup convergence of FedAvg w.r.t the number of active workers in the partial parcipation setting. The third row reports the linear speedup convergence of Nesterov accelerated FedAvg in full participation setting. 
}
\vspace{-1em}
\label{fig:speedup}
\end{figure*}

@@

In this section, we empirically examine the linear speedup convergence of FedAvg and Nesterov accelerated FedAvg in various settings, including strongly convex function, convex smooth function, and overparameterized objectives, as analyzed in previous sections.

\textbf{Setup.} Following the experimental setting in~\cite{stich2018local}, we
conduct experiments on both synthetic datasets and real-world dataset
w8a~\cite{platt1998fast}, which consists of $n = 49749 $ samples with
feature dimension $d = 300$. The label is either positive one or negative one.
The dataset has sparse binary features in $\{0, 1\}$. Each sample
has 11.15 non-zero feature values out of $300$ features on average.
We set the batch size equal to four across all experiments.
For all the results, we set random seeds as $0, 1, 2$
and report the best convergence rate across the three folds.

In general, we consider the distributed
objectives $F(\vw) = \sum_{k=1}^N p_kF_k(\vw)$, and the objective function on the
$k$-th local device includes three cases: 1) Strongly convex
objective, 2) Convex smooth objective, and 3) Overparameterized setting.
To verify the linear speedup convergence as shown in Theorems~\ref{thm:SGD_scvx}~\ref{thm:SGD_cvx}~\ref{thm:nesterov_scvx}~\ref{thm:Nesterov_cvx}, we evaluate the number of iterations needed to reach
$\epsilon$-accuracy in three objectives. We initialize all runs with $\vw_0 = \textbf{0}_d$ and measure the number of iterations to reach the target accuracy $\epsilon$. For each configuration $(E, K)$, we extensively search the learning rate from $\min(\eta_0, \frac{nc}{1 + t})$, where
$\eta_0 \in \{0.1, 0.12, 1, 32 \}$ according to different problems and $c$ can
take the values $c = 2^i \ \forall i \in \ZZ$. As the results shown in Figure~\ref{fig:speedup},
the number of iterations decreases as the number of (active) workers increasing, which is consistent for FedAvg across all three objectives under the setting of full participation (first row, Figure~\ref{fig:speedup}) and partial participation (second row, Figure~\ref{fig:speedup}). For Nesterov accelerated FedAvg, we also empirically verified the linear speedup is achieved for all three objectives under full participation.
In the next following subsections,we introduce parameter searching in for different types of objectives separately.

\subsection{Strongly Convex Objectives}
We first consider the strongly convex objective function, where we use
a regularized binary logistic regression ($F_k(\vw) =
\frac{1}{N_k} \sum_{i=1}^{N_k} \log( 1+ \exp(-y_i^k \vw^T\vx_i^k) + \frac{\lambda}{2}
\|\vw\|^2$) with regularization $\lambda=1/n\approx 2e-5$. We evenly distributed on $1, 2, 4, 8, 16, 32$ devices and  report the number of iterations/rounds needed to converge to $\epsilon-$accuracy, where $\epsilon=0.005$. The optimal objective function value $f^*$
is set as $f^* = 0.126433176216545$. This is determined numerically and we follow the setting in~\cite{stich2018local}. The learning rate is decayed as the $\eta_t = \min(\eta_0, \frac{nc}{1 + t})$, where we extensively search the best learning rate $c \in \{2^{-1}c_0, 2^{-2}c_0, c_0, 2c_0, 2^{2}c_0\}$. In this case, we search the initial learning rate $\eta_0\in \{1, 32\}$ and $c_0 = 1/8$.

\subsection{Convex Smooth Objectives}
We also use binary logistic regression without regularization.
The setting is almost same as its regularized counter part. We also evenly distributed all the samples on $1, 2, 4, 8, 16, 32$ devices. The figure shows the number of iterations needed to converge to $\epsilon-$accuracy, where $\epsilon=0.02$. The optiaml objective function value is set as $f^*=0.11379089057514849$, determined numerically. 
The learning rate is decayed as the $\eta_t = \min(\eta_0, \frac{nc}{1 + t})$, where we extensively search the best learning rate $c \in \{2^{-1}c_0, 2^{-2}c_0, c_0, 2c_0, 2^{2}c_0\}$. In this case, we search the initial learning rate $\eta_0\in \{1, 32\}$ and $c_0 = 1/8$.

\subsection{Linear regression}
We use the linear regression problem ($F_k(\vw) =
\frac{1}{N_k} \sum_{i=1}^{N_k} (\vw^T\vx_i^k + b  - y_i^k)^2$)) without adding noise to the label
for the experiments under overparameterized setting.
For linear regression, we use the same feature vectors from w8a dataset 
and generate ground truth $[\vw^*, b^*]$ from a multivariate normal distribution
with zero mean and standard deviation one. Then we generate label 
based on $y_i = \vx_i^t\vw^* + b^*$. This procedure will ensure we satisfy
the over-parameterized setting as required in our theorems. 
We also evenly distributed all the samples on $1, 2, 4, 8, 16, 32$ devices. The figure shows the number of iterations needed to converge to $\epsilon-$accuracy, where $\epsilon=0.02$. The optiaml objective function value is $f^*=0$. 
The learning rate is decayed as the $\eta_t = \min(\eta_0, \frac{nc}{1 + t})$, where we extensively search the best learning rate $c \in \{2^{-1}c_0, 2^{-2}c_0, c_0, 2c_0, 2^{2}c_0\}$. In this case, we search the initial learning rate $\eta_0\in \{0.1, 0.12\}$ and $c_0 = 1/256$.

\subsection{Partial Participation}
To examine the linear speedup of FedAvg in partial participation setting,
we evenly distributed data on $4, 8, 16, 32, 64, 128$ devices and 
uniformly sample $50\%$ devices without replacement. 
All other hyperparameters are the same as previous sections.

\subsection{Nesterov accelerated FedAvg}
The experiments of Nesterov accelerated FedAvg (the update formula is given as follows) uses the same setting as
previous three sections for vanilla FedAvg.
\begin{align*}
\vy_{t+1}^{k} & =\vw_{t}^{k}-\alpha_{t}\vg_{t,k}\\
\vw_{t+1}^{k} & =\begin{cases}
\vy_{t+1}^{k}+\beta_{t}(\vy_{t+1}^{k}-\vy_{t}^{k}) & \text{if }t+1\notin\mathcal{I}_{E}\\
\sum_{k\in\mathcal{S}_{t+1}}\left(\vy_{t+1}^{k}+\beta_{t}(\vy_{t+1}^{k}-\vy_{t}^{k})\right) & \text{if }t+1\in\mathcal{I}_{E}
\end{cases}
\end{align*}
We set $\beta_t = 0.1$ and search $\alpha_t$ in the same way as $\eta_t$
in FedAvg.

\subsection{The impact of $E$.}
In this subsection, we further examine how does the number of local steps ($E$) 
affect convergence. As shown in Figure~\ref{fig:e}, the number of iterations increases as $E$ increase, which slow down the convergence in
terms of gradient computation. However, it can save communication costs as
the number of rounds decreased when the $E$ increases. This showcase that
we need a proper choice of $E$ to trade-off the communication cost and
convergence speed. 
\begin{figure*}
\centering
	\begin{tabular}{ccc}
	\hspace{-2em} \includegraphics[width=0.33\textwidth]{fig/paper-stronglycvxsmthspeedupEpochsT-min-w8a-epsilon0131-reg1e-05.pdf} &
	\includegraphics[width=0.33\textwidth]{fig/paper-cvxsmoothspeedupEpochsT-min-w8a-epsilon0134-reg0.pdf} & 
	\includegraphics[width=0.33\textwidth]{fig/paper-linregression-newspeedupEpochsT-min-linearregressionw8a-epsilon002-reg0.pdf} \\
	\hspace{-2em} \includegraphics[width=0.33\textwidth]{fig/paper-stronglycvxsmthspeedupEpochsRounds-min-w8a-epsilon0131-reg1e-05.pdf} &
	\includegraphics[width=0.33\textwidth]{fig/paper-cvxsmoothspeedupEpochsRounds-min-w8a-epsilon0134-reg0.pdf} & 
	\includegraphics[width=0.33\textwidth]{fig/paper-linregression-newspeedupEpochsRounds-min-linearregressionw8a-epsilon002-reg0.pdf} \\
(a) Strongly convex objective & (b) Convex smooth objective & (c) Linear regression
	\end{tabular}
\caption{The convergence of FedAvg w.r.t the number of local steps $E$. }
\label{fig:e}
\end{figure*}

% ~\ref{th:strongcvxsmth}~\ref{th:cvxsmoth}~\ref{th:fedmasslr}
% \textbf{Linear speedup of FedAvg and Nesterov accelerated FedAvg.}

@@
